---
title: "new_study_MI"
output: pdf_document
date: "2024-02-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multiple imputation of new study

## EDA

### Loading packages and data

```{r}
library("readxl")
library("dplyr")
library("glue")
library("mice")
library("survival")
library("SurvMetrics")
library("ggplot2")
library("ggrepel")
library("caret")
library("reshape2")
```

```{r cars}
path_erlend = "../../data/GEP-NEN-data/updated/new_study.xlsx"
path_nina = "../../Both/new_study.xlsx"
df.raw = read_excel(path_nina, sheet = 1, na = "NaN")
head(df.raw)
```


### Double check low variance columns

```{r}
print_low_variance_cols <- function(df, threshold = 0.9) {
  # This function will print the names of columns with low variance along with their value counts.
  # For numeric columns, it calculates the variance.
  # For categorical columns, it checks if the proportion of the most common value is above the threshold (default is 90%),
  # and prints the counts of each unique value.
  
  for (col_name in names(df)) {
    col <- df[[col_name]]
    
    if (is.numeric(col)) {
      # If the column is numeric, calculate variance.
      var_col <- var(col, na.rm = TRUE)
      if (var_col < threshold) {
        cat(col_name, "has low variance:", var_col, "\n")
      }
    } else {
      # If the column is not numeric, check the proportion of the most common value.
      freq <- table(col)
      prop_max <- max(freq) / sum(freq)
      if (prop_max > threshold) {
        cat(col_name, "has low variance with predominant value: ", names(freq)[which.max(freq)], "\n")
        cat("Value counts: ", paste(names(freq), "=", freq, collapse=", "), "\n\n")
      }
    }
  }
}

print_low_variance_cols(df.raw)

```


### Removing features and samples (mostly done)

```{r}
# Not used, also linear dependence
to_remove_1 = c("Ki-67 grouping", "OS (months)")

# remove PATNO, Age at death keep?
to_remove_2 = c("PATNO", "Age at Death")

df <- subset(df.raw, select = !(names(df.raw) %in% to_remove_1))
df <- subset(df, select = !(names(df) %in% to_remove_2))
```

### Investigating linear relationships

```{r}
# Function to encode variables
encode_to_numeric <- function(df) {
  df[] <- lapply(df, function(x) {
    if(is.factor(x) | is.character(x)) {
      x <- as.factor(x) # Convert to factor if it's not already
      return(as.numeric(x)) # Convert factor to numeric
    } else {
      return(x) # Keep as is if numeric
    }
  })
  return(df)
}

df_numeric <- encode_to_numeric(df)
```

```{r}
corr = cor(df_numeric, use = "pairwise.complete.obs")
```
```{r}
cor_matrix <- corr
# Set NA to 0 
cor_matrix[is.na(cor_matrix)] <- 0

cor_above_threshold <- cor_matrix > 0.5
diag(cor_above_threshold) <- FALSE
keep_vars <- apply(cor_above_threshold, 1, any)

filtered_cor_matrix <- cor_matrix[keep_vars, keep_vars]

melted_filtered_matrix = melt(filtered_cor_matrix)
g = ggplot(melted_filtered_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5, limit = c(0.3, 1),
                       space = "Lab", name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        axis.title = element_blank()) +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(title = "Correlation Matrix Heatmap (Filtered > 0.5)")

# Optionally, save the plot with increased dimensions for a larger view
ggsave("filtered_heatmap_plot.png", plot = g, width = 20, height = 15, dpi = 300)
```

We can definitely see some strongly (and also perfectly) correlated features

```{r}
# Remove highly correlated features
to_remove_3 = c("M-stage", "TNM-staging", "Time from diag to mets (months)")
df <- subset(df, select = !(names(df) %in% to_remove_3))

```

### Encoding

```{r}
# Removed Metastatic disease (from correlation)
one_hot_enc = c("Primary Tumour",
                "Primary Tumour Resected", 
                "Prior Other Cancer",
                "Sex",
                "Tumour Morphology",
                "Haemoglobin",
                "Platelets",
                "WBC",
                "Mets(Bone)",
              #  "Mets(Brain)", removed because low variance
                "Mets(LN)",
                "Mets(Liver)",
                "Mets(Lung)", 
                "Mets(Other)",
                "Chemotherapy Type",
              #  "Reintroduction with Cisplatin+Etoposide",
                "Co-morbidity Severity",
               # "TNM-staging", removed because > 90 corr
              #  "M-stage",
                #"Radical Surgery",
                #"FDG-PET Uptake", removed because constant
                #"NEC/MANEC",
                "Co-morbidity",
                "Hist Exam Metastasis",
                "Hist Exam Primary Tumour",
                #"Treatment Intention",
                "Dev of Bone Mets",
                #"Dev of Brain Mets",
                "Loc Adv Resectable Disease",
               # "Mets(Skin)", low variation
                "Mets(LN Retro)",
                "Mets(LN Regional)", 
                "Mets(LN Distant)",
                "Treatment Stopped",
                # "Bone Mets", removed because collinearity
              #  "Metastatic Disease", removed because low variation
                "Living Alone",
                #"Creatinine", 
                "Stage grouped",
                "Smoking",
                "Best Response (RECIST)")

# check values
for (col in one_hot_enc){
  print(glue("Column: {col}, values: {unique(df[col])}\n"))
}
```

Co-morbidity and co-morbidity severity may be MAR (severity missing)

```{r}
binary = c("Primary Tumour Resected", 
           "Prior Other Cancer", 
           "Sex", 
           "Haemoglobin", 
           "Platelets",
           "WBC",
           "Mets(Bone)",
           "Mets(LN)",
           "Mets(Liver)",
           "Mets(Lung)",
           "Mets(Other)",
           "Co-morbidity Severity",
           #"Radical Surgery",
           #"NEC/MANEC",
           "Co-morbidity",
           "Hist Exam Metastasis",
           "Hist Exam Primary Tumour",
           #"Treatment Intention",
           "Dev of Bone Mets",
           #"Dev of Brain Mets",
           "Loc Adv Resectable Disease",
           "Mets(LN Retro)",
           "Mets(LN Regional)",
           "Mets(LN Distant)",
           "Living Alone",
           #"Creatinine",
           "Stage grouped")


factor_gr_2 = c("Primary Tumour",
                "Tumour Morphology",
                "Chemotherapy Type", 
                "Treatment Stopped",
                "Smoking",
                "Best Response (RECIST)")

combined = c(binary, factor_gr_2)
all(one_hot_enc %in% combined)
```

Not sure if encoding is correct in R? Henning suggested order 0, 1, 2 for Negative, Partly Positive and Strongly Positive.

```{r}
factor_levels <- list(
  `Chromogranin A2` = c("Normal", ">Normal <= 2UNL", "> 2UNL"), 
  Differentiation = c("Highly Differentiated", "Intermediate", "Poorly Differentiated"),
  `Chromogranin A` = c("Negative", "Partly Positive", "Strongly Positive"),
  NSE = c("Normal", ">Normal <= 2UNL", "> 2UNL"),
  Synaptophysin = c("Negative", "Partly Positive", "Strongly Positive"),
  `WHO Perf Stat` = c("WHO 0", "WHO 1", "WHO 2", "WHO 3", "WHO 4"),
  ALP = c("Normal", ">Normal <= 3 UNL", ">3 UNL"),
  LDH = c("Normal", ">Normal <= 2UNL", "> 2UNL"),
  #`CD-56` = c("Negative", "Partly Positive", "Strongly Positive"), Removed?
  `T-stage` = c("Tx", "T0", "T2", "T3", "T4"),
  `N-stage` = c("Nx", "N0", "N1", ">N1"),
  #SRI = c("Negative", "< Liver", "> Liver"),
  Octreoscan = c("Negative", "< Liver", "> Liver"),
  `CD-56` = c('Negative', 'Partly Positive', 'Strongly Positive') 
)
```

```{r}
accounted_for = c(names(factor_levels), factor_gr_2, binary)
df[,names(df)[!names(df) %in% accounted_for]]
```

Manually encode the rest

```{r}
integer = c("BMI",
            "Ki-67",
            "Absolute Neutrophil Count",
            "Albumin",
            "CRP",
            "Number of Courses",
            "OS (days)",
            "Time from diag to mets (days)",
            "Age at Diagnosis")

# Try numeric instead
numeric = c("BMI",
            "Ki-67",
            "Absolute Neutrophil Count",
            "Albumin",
            "CRP",
            "Number of Courses",
            "OS (days)",
            "Time from diag to mets (days)",
            "Age at Diagnosis")

bool = c("Status")

#df[integer] = lapply(df[integer], as.integer)
df[numeric] = lapply(df[numeric], as.numeric)
df$Status = ifelse(df$Status == "Dead", 1, 0)
df[bool] = lapply(df[bool], as.logical)

df[binary] = lapply(df[binary], factor)
df[factor_gr_2] = lapply(df[factor_gr_2], factor, ordered = FALSE)
df[names(factor_levels)] <- lapply(names(factor_levels), function(col_name) {
  factor(df[[col_name]], levels = factor_levels[[col_name]])
})
```

### Changing column names (technicality in R)

```{r}
rename_col = function(old, new, df){
  names(df)[names(df) == old] <- new
  return (df)
}

fixColNames <- function(df) {
  # Use gsub to replace spaces with underscores in each column name
  names(df) <- gsub(" ", "_", names(df))
  names(df) <- gsub("\\+", "Plus", names(df))
  names(df) <- gsub("\\-", "Minus", names(df))
  names(df) <- gsub("\\.", "Dot", names(df))
  names(df) <- gsub("\\(", "leftB", names(df))
  names(df) <- gsub("\\)", "rightB", names(df))
  names(df) <- gsub("\\__", "double_under", names(df))
  names(df) <- gsub("/", "slash", names(df))
  names(df) <- gsub("1", "uno", names(df))
  names(df) <- gsub("2", "two", names(df))
  names(df) <- gsub("3", "three", names(df))
  names(df) <- gsub("56", "fiftysix", names(df))
  names(df) <- gsub("67", "sixtyseven", names(df))

  # Return the modified dataframe
  return(df)
}

df.renamed = fixColNames(df)
```

```{r}
print(str(df.renamed))
```

## Multiple Imputation

```{r}
df.mice = df.renamed %>% subset(select = -c(Status, OS_leftBdaysrightB))
```

### Initial (dry run)

Dry run to investigate correct setup

```{r}
ini = mice(df.mice, maxit = 0)
ini
```


```{r}
ini$nmis
```

Extract the method to alter the ordered ones (they are currently treated as polytomous)

```{r}
me = ini$method
me
```

```{r}
names(me)
```

```{r}
names(factor_levels)
```

```{r}
ordered = c("Chromogranin_Atwo",
            "Differentiation", 
            "Chromogranin_A",
            "NSE", 
            "Synaptophysin",
            "WHO_Perf_Stat", 
            "ALP",
            "LDH",
    #        "CDMinusfiftysix",
            "TMinusstage",
            "NMinusstage")
  
for (col in ordered){
  me[col] = "polr"
}

me
```

### Imputation by PMM


```{r}
G = 50

imp = mice(df.mice, m = G, pred = predictor_matrix, me = "pmm", seed = 124124)
```

```{r}
imp$loggedEvents
```

### Imputation by specifying the models

```{r}
predictor_matrix = quickpred(df.mice, mincor = 0.1)
```


```{r}
imp = mice(df.mice, m = G, pred = predictor_matrix, me = me, seed = 124124)
```


```{r}
imp$loggedEvents
```


### Assessing convergence

```{r}
plot(imp)
```

```{r}
complete(imp, 1)
```

## Saving the imputed datasets

```{r}
revert_ColNames <- function(df) {
  # Reverse replacements made by fixColNames
  names(df) <- gsub("sixtyseven", "67", names(df))
  names(df) <- gsub("fiftysix", "56", names(df))
  names(df) <- gsub("three", "3", names(df))
  names(df) <- gsub("two", "2", names(df))
  names(df) <- gsub("uno", "1", names(df))
  names(df) <- gsub("slash", "/", names(df))
  names(df) <- gsub("double_under", "__", names(df))
  names(df) <- gsub("rightB", ")", names(df))
  names(df) <- gsub("leftB", "(", names(df))
  names(df) <- gsub("Dot", ".", names(df))
  names(df) <- gsub("Minus", "-", names(df))
  names(df) <- gsub("Plus", "+", names(df))
  names(df) <- gsub("_", " ", names(df)) # This replacement is done last to avoid conflicts

  return(df)
}

# INSERT PATH NINA
file_path_nina = "../R/datasets/pmm/"
#file_path_nina = "/Users/ninarebeccalizana/Documents/Master/code/New-Study/R/datasets/single"
file_path_erlend = "../../data/R/MI/new_study/"

save_datasets = function(mice.obj, file_path){
  M = mice.obj$m
  for (m in seq(1, M)){
    df.m = complete(mice.obj, m)
    df.m = revert_ColNames(df.m)
    df.m[, c("status", "time", "PATNO")] = df.raw[, c("Status", "OS (days)", "PATNO")]

    write.csv(df.m, glue("{file_path}new_studyM{m}.csv"))
    }
  }
```

```{r}
#save_datasets(mice.obj = imp, file_path = file_path_erlend)
save_datasets(mice.obj = imp, file_path = file_path_nina)
```

